Orchestrating stencils with the Multi-Stencil Framework

### 2017-02-17 ARTICLES/IJPP/articleR1/formalism.tex
restructuration du plan + figure MSF
nouvelle partie II + ajout partie IV

@@

This section gives a complete formal description of what we call a \textit{stencil program} and its computations.
The proposed language MSL (Multi-Stencil Language) does not depend on the type of space discretization used.
Thus, some details given in the following sections, are useful to understand the domain but do not have to be kept in mind to understand the Multi-Stencil Language.
% As a result, at the end of each subsection a summary of what is useful to understand the language is given.

@@

\begin{mydef}
A \textit{multi-stencil program} is defined by the octuplet 
\begin{equation*}
\mathcal{MSP}(\mathcal{M},\Phi,\mathcal{D},\mathcal{N},\Delta, \mathcal{S},T,\Gamma)
\end{equation*}
\end{mydef}

@@

\medskip
A stencil program has been formally defined in this section. This formalism is used in the next Section to define two parallelization techniques of a multi-stencil program.

One can note that all definitions given in this section are independent from the topology of the mesh. This property will be kept in the rest of this paper to propose the mesh-agnostic MSL language.

>>>>

\HC{has to be revised according to modifications + see the scatter problem mentioned by reviewers}

@@

This section gives a complete formal description of what we call a \textit{stencil program} and its computations. For this formal description is extracted a Multi-Stencil Program meta-model which will be used to define MSL and GA in the next section.
%The proposed language MSL (Multi-Stencil Language) does not depend on the type of space discretization used.
%Thus, some details given in the following sections, are useful to understand the domain but do not have to be kept in mind to understand the Multi-Stencil Language.

@@

\begin{mydef}
A \textit{multi-stencil program} is defined by the octuplet 
\begin{equation}
\mathcal{MSP}(\mathcal{M},\Phi,\mathcal{D},\mathcal{N},\Delta, \mathcal{S},T,\Gamma)
\label{eq:msp}
\end{equation}
\end{mydef}

@@

\medskip
A stencil program has been formally defined in this section. This formalism is used in the next Section as a meta-model to define both a the domain specific language MSL, and the generic assembly of a multi-stencil program GA.

One can note that all definitions given in this section are independent from the topology of the mesh. This property will be kept in the rest of this paper so that MSF has the capacity to be used for any kind of mesh, thus being mesh-agnostic.

### 2017-02-17 ARTICLES/IJPP/articleR1/msf.tex
restructuration du plan + figure MSF
nouvelle partie II + ajout partie IV

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% présenter le framework dans sa vision d'ensemble comme dans WOLFHPC
% faire un joli schéma sous inkscape
% éléments à représenter et expliquer :
%	- MSL language
%	- generic component assembly
%	- MSC compiler
%	- final component assembly
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------
\subsection{Background on component models}
%----------------------------------------
Component-based software engineering (CBSE) is a domain of software engineering~\cite{Szyperski:2002:CSB:515228} which promotes code
re-use, separations of concerns, and thus maintainability.
An application is made of a set of component instances.
A component is a black box that implements an independent functionality of the application, and which interacts with its environment only through well defined interfaces: its ports.
A port can, for example, specify services provided or required by the component.
With respect to high performance computing, some works have also shown
that component models can achieve the needed level of performance, and
scalability while also helping in application portability~\cite{Bernholdt01052006, bigot:inria-00388508, UCHPC2015}

@@

\llc~\cite{l2c} is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class.
The services offered by the components are specified trough $provide$ ports,
those used either by $use$ ports for a single service instance,
or $use-multiple$ ports for multiple service instances.
Services are specified as \texttt{C++} interfaces.
\llc also offers $MPI$ ports that enable components to share MPI communicators.
Components can also have attribute ports to be configured.
%
As illustrated in Figures~\ref{fig:ports}, a $provide$ port is
represented by a white circle, a $use$
port with a black circle, a $use-multiple$ port by a black circle with
a white $m$ in it. MPI port are
connected with a black rectangle.
A \llc-based application is a static \emph{assembly} of components instances and connections between their ports.
Such an assembly is described in LAD, an XML dialect.

@@

%----------------------------------------
\subsection{Multi-Stencil Framework overview}
%----------------------------------------

@@

>>>>

\HC{under construction, present the framework}

This section first presents a background on component models and particularly on the Low Level Components. This background is needed to understand the second part of the section which presents an overview of the overall Multi-Stencil Framework (MSF).

%----------------------------------------
\subsection{Background on component models}
%----------------------------------------
Component-based software engineering (CBSE) is a domain of software engineering~\cite{Szyperski:2002:CSB:515228} which improves code re-use, separations of concerns, and thus maintainability.
An component-based application is made of a set of component instances linked together, this is also called a component \emph{assembly}.
A component is a black box that implements an independent functionality of the application, and which interacts with its environment only through well defined interfaces: its ports.
A port can, for example, specify services provided or required by the component.
With respect to high performance computing, some works have also shown
that component models can achieve the needed level of performance, and
scalability while also helping in application portability~\cite{Bernholdt01052006, bigot:inria-00388508, UCHPC2015}

@@

\llc~\cite{l2c} is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class.
The services offered by the components are specified trough $provide$ ports,
those used either by $use$ ports for a single service instance,
or $use-multiple$ ports for multiple services instances.
Services are specified as \texttt{C++} interfaces.
\llc also offers $MPI$ ports that enable components to share MPI communicators. Finally, 
components can also have attribute ports to be configured.
%
In this paper, and as illustrated in Figures~\ref{fig:ports}, a $provide$ port is represented by a white circle, a $use$ port with a black circle, a $use-multiple$ port by a black circle with a white $m$ in it. MPI port are connected with a black rectangle. A \llc-based application is a static \emph{assembly} of components instances and the connections between their ports.
Such an assembly is described in LAD, an XML dialect.

@@

%----------------------------------------
\subsection{Multi-Stencil Framework overview}
%----------------------------------------
The Multi-Stencil Framework helps end-users to produce high performance parallel applications for the specific case of multi-stencils. The multi-stencil domain will be formally defined in the next section. A multi-stencil program is typically a numerical simulation based on explicit numerical schemes that approximate PDEs equations with one or multiple neighborhood computations.

Figure~\ref{fig:msf} presents the overview of the Multi-Stencil Framework that will be entirely detailed throughout this paper. It is composed of four different parts presented below.

One can note in this figure that MSF is addressed to two different kinds of end-users: the \emph{numerician}, in other words the mathematician, and the \emph{developer}. Most of the time numericians do have programming knowledge, however as it is not their core domain and because of a lack of time, development is often left to engineers according to numericians needs. MSF has the interesting particularity to propose a clear separation of concerns between these two end-users by distinguishing the description language from the numerical code.

MSF also has the interesting capability to become flexible thanks to a third user possibility to interact with the framework: the High Performance Computing (HPC) specialist.

@@

\paragraph{\textbf{Multi-Stencil Language}}
The Multi-Stencil Language, or MSL, is the domain specific language proposed by the framework for the numerician. It is a descriptive language, easy to use, without any concern onto implementation details, which fits the need of a mathematician to describe the simulation. The description written with MSL can be considered as an input of the framework. MSL is described in details in Section~\ref{sect:msl}. The language is built upon the meta-model described in Section~\ref{sect:formalism}.

\paragraph{\textbf{Generic Assembly}}
In addition to the language MSL and the simulation description written by the numerician, MSF needs a generic assembly of a multi-stencil program as input. What is called a generic assembly is a component assembly for which meta-types are represented and for which some parts need to be generated or specialized. It is similar to a generic pattern or a skeleton in usual programming models such as C++ or functional models. From this generic assembly will be built the final specialized assembly of the simulation. As well as MSL, this generic assembly is described in Section~\ref{sect:msl} and is built upon the meta-model described in Section~\ref{sect:formalism}. Moreover, an example of generic assembly is used in evaluations of Section~\ref{sect:eval}.
\HC{Note idée : HPC specialist gives types for DDS and Data}

\paragraph{\textbf{Multi-Stencil Compiler}}
The core of the framework is the Multi-Stencil Compiler, or MSC. It is responsible for transforming the genric assembly into the final parallel assembly which is specific to the simulation described by the numerician. MSC is described in Sections~\ref{sect:parallelism} and~\ref{sect:msp}.

\paragraph{\textbf{Specialized assembly}}
Finally, the input of MSF is the component assembly generated by MSC. It is an instanciation and a transformation of the generic component assembly, by adding component types and by adding specific components generated by MSC. From this final component assembly which is specific to the simulation initially described by using MSL, the developer will finally write numerical codes.

### 2017-02-17 ARTICLES/IJPP/articleR1/msl.tex
restructuration du plan + figure MSF
nouvelle partie II + ajout partie IV

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% unifier texte avec nouvelle version article
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%----------------------------------------
\subsection{Generic Assembly}
%----------------------------------------

@@

   \node[provide,right=0.15 of Tu] (Cp) {};
   \node[component, draw=green!80!black,right=2cm of Cp] (C) {$Computations$};
   \node[use,above=0.8cm of C] (Cu) {};
   \node[right=0.2cm of Cu] (star) {$*$};
   \node[right=1.5cm of C] (gamma) {$\Gamma$};

   \node[provide,below right=0.2 of Du3] (Datap1) {};
   \node[provide,above=0.15 of Cu] (Datap2) {};
   \node[component,double, draw=red,above=0.8cm of Datap2] (Data) {$Data$};
   \node[use,above=0.8cm of Data] (Datau) {};
   \node[left=1cm of Data] (delta) {$\Delta,\mathcal{D}$};
   \node[mpi,right=1cm of Data] (mpidata) {};
   \node[above=0.2cm of mpidata] (star2) {$*$};

   \node[provide,right=0.15 of Du1] (DDSp1) {};
   \node[provide,above=0.15 of Datau] (DDSp2) {};
   \node[component, draw=red,above=0.8cm of DDSp2] (DDS) {$DDS$};
   \node[right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{E}$};

@@

@@

From the formalism detailed in the previous section, the Multi-Stencil Language and its grammar can already be given.
This grammar is sufficient to automatically extract parallelism from the program as will be explained in the next section.

@@

The grammar of the Multi-Stencil Language is given in Figure~\ref{fig:grammar} and an example is provided in Figure~\ref{fig:exmsl}.
A Multi-Stencil program is composed of eight parts.

>>>>

\HC{under revision / has to be revised according to modifications}

%----------------------------------------
\subsection{Generic Assembly}
%----------------------------------------
From the meta-model represented by Equation~\ref{eq:msp} the generic assembly of any multi-stencil program can be built. It is represented by Figure~\ref{fig:ga}. As indicated, the GA has five different components described below.

@@

   \node[provide,right=0.15 of Tu] (Cp) {};
   \node[component, draw=green!80!black, text=blue!80!black,right=2cm of Cp] (C) {$Computations$};
   \node[use,above=0.8cm of C] (Cu) {};
   \node[right=0.2cm of Cu] (star) {$*$};
   \node[right=1.5cm of C] (gamma) {$\Gamma$};

   \node[provide,below right=0.2 of Du3] (Datap1) {};
   \node[provide,above=0.15 of Cu] (Datap2) {};
   \node[component,double, draw=red,above=0.8cm of Datap2] (Data) {$Data$};
   \node[use,above=0.8cm of Data] (Datau) {};
   \node[left=1cm of Data] (delta) {$\Delta$};
   \node[mpi,right=1cm of Data] (mpidata) {};
   \node[above=0.2cm of mpidata] (star2) {$*$};

   \node[provide,right=0.15 of Du1] (DDSp1) {};
   \node[provide,above=0.15 of Datau] (DDSp2) {};
   \node[component, draw=red,above=0.8cm of DDSp2] (DDS) {$DDS$};
   \node[right=1.5cm of DDS] (m) {$\mathcal{M},\Phi,\mathcal{D},\mathcal{N}$};

@@

\paragraph{\textbf{Driver}}
This component can be compared to the main function of a usual program. It is responsible for initialization and execution of other components, just like variable initializations and function calls.

\paragraph{\textbf{Time}}
As indicated in the figure, this component is responsible for the time $T$ defined in Equation~\ref{eq:msp}. Thus it is composed of a time loop and eventually of a convergence reduction.

\paragraph{\textbf{DDS}}
This component is responsible for the mesh and its entities $\mathcal{M}$, and $\Phi$, the computation domains $\mathcal{D}$ and the set of neighborhood functions $\mathcal{N}$. Moreover, if a data parallelization is needed, this component should handle mesh partitioning and should provide a synchronization interface as it will be detailed in Section~\ref{sect:parallelism}. One can note that this component is red, this means that this component should be provided by a HPC specialist in MSF as illustrated in the previous Figure~\ref{fig:msf}.

\paragraph{\textbf{Data}}
This component is inseparable from the previous one $DDS$. It actually uses the DDS component to instanciate one quantity. Thus, it is responsible for the set of quantities $\Delta$. This component is also red. As $DDS$ it should be provided by a HPC specialist. Moreover, it is represented by a double line box which means that it can be instanciated multiple time will others will be instanciated once. Actually, this component will be instanciated as many times as the number of quantities in the simulation.

\paragraph{\textbf{Computations}}
Finally, this component is responsible for $\Gamma$, \ie the computations of the simulation. This component is green and blue because it will be partly generated by MSC and partly feeded by the developer. Actually, this component will be replaced by MSC by a sub-assembly of component for which the parallel part will be automatically generated, and for which components responsible for simulation code will be completed by the developer.

@@

The second element of MSF which is built upon the meta-model represented by Equation~\ref{eq:msp} is the Multi-Stencil Language and its grammar. This grammar is sufficient (in addition to GA) for MSC to automatically extract a parallel pattern of the simulation which is finally dumped as a specialized instanciation of GA.

@@

The grammar of the Multi-Stencil Language is given in Figure~\ref{fig:grammar} and an example is provided in Figure~\ref{fig:exmsl}. A Multi-Stencil program is composed of eight parts as in Equation~ref{eq:msp}.
